Entrainment of low Mach number thermals in stratified domains

### 2019-05-02 tex/stratified_thermals.tex
writes intro; rearranges and improves experiment section.

%%%%% Body of the paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% INTRODUCTION
\section{Introduction}
\label{sec:intro}
(Para 1) Brief intro to solar convective conundrum.

(Para 2) Entropy rain history and expansion by Axel

(Para 3) Boussinesq thermals and their filling factor vs. depth.

(Para 4) What we're doing here, how it fits into this context.

(Para 5) Standard last paragraph of an intro?

@@

\subsection{Phenomenological description of thermal evolution}
The initial conditions of the thermals studied in this work are spherical
entropy perturbations with zero initial velocity, as depicted in 
Fig.~\ref{fig:evolution_colormeshes}a. Buoyant forces cause these perturbations
to fall and develop into vortex rings, as depicted in 
Figs.~\ref{fig:evolution_colormeshes}c-f.

\begin{figure*}[t!]
    \includegraphics[width=\textwidth]{evolution_colormeshes.png}
    \caption{
    \label{fig:evolution_colormeshes} }
\end{figure*}

@@

We further assume that the volume-integrated momentum, $P_z \approx \rho V w_{th}$,
where $w_{th}$ is the vertical velocity of the thermal as a whole, and the volume
of the fluid region propogating with the thermal is a spheroid, $V = V_0 r^3$, for
some constant $V_0$. Plugging Eqn.~\ref{eqn:r_theory} into this approximation, we
find that
\begin{equation}
\rho^{-1/2}w = \left(\frac{(\pi\Gamma)^{3/2}}{V_0}\right)\frac{\beta B t + M_0}{(B t + I_0)^{3/2}}.
\end{equation}
Substituting $w = dz/dt$ and $\tau \equiv (Bt + I_0)/\Gamma$, We retrieve
\begin{equation}
\frac{dz}{\sqrt{\rho(z)}} 
= \left(\frac{\pi^{3/2}\Gamma}{V_0}\right)
\left(\beta\Gamma\tau^{-1/2} + [M_0 - \beta I_0]\tau^{-3/2}\right)d\tau.
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this
result can be straightforwardly integrated with $\rho(z)$ plugged in in order to
find the position of the thermal versus time. While we will leave this result
general for now, we will plug in our polytropic stratification at the end of
section \ref{sec:experiment}, and show that the resulting expression does a
good job of explaining the evolution of thermals in these atmospheres in
section \ref{sec:results}.

@@

\subsection{Polytropic Atmospheres}
We study an ideal gas whose equation of state is $P = \rho T$ and whose stratification
is a perfectly adiabatic polytrope,
\begin{gather}
T_0 = 1 + (\grad_{\text{ad}})(z - L_z) \\
\rho_0 = T_0^{m_{\text{ad}}},
\end{gather}
where $m_{\text{ad}} = (\gamma-1)^{-1}$, and where the adiabatic 
temperature gradient is $\grad_{\text{ad}} = - g / c_P$.

\subsection{Anelastic Equations}
The anelastic equations are \citep{lecoanet&all2014},
\begin{gather}
\td{\grad}\cdot\td{\bm{u}} = -\td{w}\partial_{\tilde{z}} \ln\rho_0 \\
\frac{D \td{\bm{u}}}{D \td{t}} = -\td{\grad} \td{\varpi} + g\frac{\td{S_1}}{c_P}\hat{z} + \frac{1}{\rho_0}\td{\grad}\cdot\left(\mu\td{\lilstressT}\right) \\
\frac{D \td{S_1}}{D\td{t}} = \frac{1}{\rho c_P}\td{\grad}\cdot\left(\kappa T_0 \td{\grad} \td{S_1}\right) + \frac{\mu}{\rho_0 T_0}\td{\sigma_{ij}}\partial_{\td{x_i}}\td{u_j}.
\end{gather}
where $D/D\td{t} = \partial/\partial \td{t} + \td{\bm{u}}\cdot\td{\grad}$ and the viscous stress tensor is
\begin{equation}
\td{\sigma_{ij}} = \left(\partial_{\td{x_i}}\td{u_j} + \partial_{\td{x_j}}\td{u_i} - \frac{2}{3}\delta_{ij}\td{\grad}\cdot\td{\bm{u}}\right).
\end{equation}

We nondimensionalize these equations as in \citet{lecoanet&jeevanjee2018} such that
the length scale is the diameter of the thermal and the velocity scale is the freefall
velocity,
\begin{equation}
\begin{split}
\td{\grad}\rightarrow(\td{L}_{th}^{-1})\grad, \qquad&
\td{S}_1 \rightarrow(\Delta\td{S})S_1,\\
\td{\bm{u}} \rightarrow (\td{u}_{th})\bm{u}, \qquad&
\td{\varpi} \rightarrow (\td{u}_{th}^2)\varpi,\\
\partial_{\tilde{t}} \rightarrow (\td{u}_{th}/\td{L}_{th})\partial_t,\qquad&
\end{split}
\end{equation}
with
\begin{equation}
\tilde{u}_{th}^2 = \frac{g \tilde{L}_{th} \Delta \tilde{s}}{c_P}, \qquad
\text{Re}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\nu}, \qquad
\text{Pr}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\chi}.
\end{equation}
The resulting equations are very similar to the boussinesq equations,
\begin{gather}
\DivU = -w \partial_z \ln\rho_0, \\
\begin{split}
\partial_t \bm{u} + \bm{u}\cdot\grad\bm{u} = \\
- \grad \varpi + S_1\hat{z} 
+ \frac{1}{\text{Re}_{\text{ff}}}\left[\grad^2 \bm{u} + \frac{1}{3}\grad(\DivU)\right] 
\end{split}\\
\begin{split}
\partial_t S_1 + \bm{u}\cdot\grad S_1 = \\
\frac{1}{\text{Re}_{\text{ff}}}\left(\frac{1}{\text{Pr}_{\text{ff}}\rho_0c_P }[\grad^2 S_1 + \partial_z\ln T_0 \cdot\partial_z S_1]\right.\\\left.
+ \frac{-(\grad_{ad})}{\rho_0 T_0}\sigma_{ij}\partial_{x_i}u_j \right)
\end{split}
\end{gather}

\subsection{Fully Compressible Equations}
The fully compressible equations, under the same assumptions as above, are
\begin{gather}
\frac{D \ln\rho}{Dt} + \DivU = 0 \\
\frac{D \bm{u}}{D t} = -\grad T - T\grad\ln\rho - g\hat{z} + \frac{1}{\rho}\grad\cdot\left(\mu\lilstressT\right) \\
\frac{D T}{Dt} + (\gamma-1)T\DivU  = \frac{1}{\rho c_V}\grad\cdot\left(\kappa \grad T\right) + \frac{\mu}{\rho c_V}\sigma_{ij}\partial_{x_i}u_j.
\end{gather}

\subsection{Initial Conditions}
We set the initial specific entropy perturbation as
\begin{equation}
S_1 = - A(1 - \text{erf}\left(\frac{r - r_{th}}{\delta}\right))/2,
\end{equation}
where in the FC cases $A = 10^{-4}$ and in the nondimensionalized AN cases,
$A = 1$. Here, $r = \sqrt{x^2 + y^2 + (z - z_0)^2}$, where $z_0 = Lz - 3*r_{th}$,
and $r_{th} = Lz/40$ is the radius of the thermal.

In the fully compressible equations, it is essential that the initial conditions be in pressure
equilibrium such that the low mach number nature of the thermal can be studied. In order to 
achieve this pressure balance, we set
\begin{equation}
\ln\rho_1 = S_1/c_P, \qquad T_1 = T_0(e^{-\ln\rho_1} - 1).
\end{equation}

\subsection{Solution for thermal evolution in a Polytrope}

>>>>

%%%%% Body of the paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% INTRODUCTION
\section{Introduction}
\label{sec:intro}
Recent observations of solar convection have revealed a convective conundrum.
Power spectra have revealed weaker flows than anticipated at large length scales,
calling into question the existence of so-called ``giant cells'' driven by deep
convection which would manifest as powerful, large-scale motions at the solar
surface. This discrepancy between theory and observations has called into question
our fundamental understanding of convection, sparking fruitful investigations into
the nature of convection in the Sun.

\cite{spruit1997} hypothesized that convective motions may be driven entirely by
cool downflows at the surface of the Sun, and \citet{brandenburg2016} expanded
upon this ``entropy rain'' hypothesis. Brandenburg's work includes a careful
expansion of mixing length theory to incorporate flux contributions from nonlocal
convective motions, and handles this theory in a horizontally-averaged sense.
He includes some discussion of possible flow morphologies which could be manifestations
of this entropy rain, and even includes some brief simulations of propogating
Hill vortices. However, these simulations and discussions did not include 
include a fundamental piece of entropy rain: it is buoyant, and has an entropy
deviation from the background atmosphere.

If entropy rain does evolve into downward propagating buoyant vortex rings, it is
important to understand how buoyancy effects the filling factor of these basic
convective elements. In the context of Earth's atmosphere, so-called ``thermals''
are thought to be the nucleus of cloud formation. Thermals are buoyant areas
of fluid which evolve into propagating buoyant vortex rings, and their
evolution in the Boussinesq limit have been well studied in the laboratory
for decades (see e.g., CITE), and more recently have been studied through
Direct Numerical Simulation (DNS) in the laminar and turbulent regime
\citep{lecoanet&jeevanjee2018}. One fundamental result of these studies of
thermals is that they experience a large degree of entrainment: their size expands
with height and their propagation velocity slows despite their buoyant nature.
However, we do not know of a study in which the propagation of these thermals,
and thus the nature of their entrainment, is affected by a 
significant atmospheric stratification.

In the absence of buoyantly-induced entrainment, \citet{brandenburg2016}
suggests that the filling factor, $f$, of vortex rings should decrease like
$f \propto \rho^{\alpha}$ for horizontal compression and $f \propto \rho^{\alpha}$ for 
spherical compression. On the other hand, the filling factor of Boussinesq
thermals \emph{increases} like $f \propto d$, where $d$ is the depth propagated.
In this work, we extend the study of \citet{lecoanet&jeevanjee2018} to study
the propagation of low-Mach number, cold thermals in stratified domains, and how
buoyant entrainment affects the scaling of their filling factor with depth. If
buoyant entrainment is a dominant effect, it is possible that these buoyant vortex rings
would simply grow too large and stall before reaching the bottom of the solar convection zone.
On the other hand, if the compression effects suggested by \citet{brandenburg2016} are the
dominant effect, then it is possible that these thermals could propagate to the bottom of the solar
convection zone, or potentially shrink to a sufficiently small size where thermal dissipation is
significant.

We lay out our work as follows. In section \ref{sec:theory}, we develop a theoretical description
of thermals in a stratified domain. In section \ref{sec:experiment}, we describe and verify
the experiments conducted in this work. In section \ref{sec:results}, we compare the results
of our experiments to the theory developed in section \ref{sec:theory}. Finally, in section
\ref{sec:discussion}, we discuss what our results imply for the entropy rain hypothesis.

@@

\subsection{Phenomenological description of thermal evolution}
We show pictorally the evolution of a cold thermal from rest in 
Fig.~\ref{fig:evolution_colormeshes}. In Fig.~\ref{fig:evolution_colormeshes}a,
the evolution of a thermal in a weakly stratified domain with
$n_\rho = 0.5$ density scale heights is shown. In Fig.~\ref{fig:evolution_colormeshes}b,
the evolution of a thermal in an appreciably stratified domain with
$n_\rho = 3$ density scale heights is shown. While the initial conditions are
identical in both domains (spherical specific entropy perturbations of the
same magnitude whose diameters are 5\% of the domain depth), and while in both cases
the thermal quickly evolves into a propagating buoyant vortex ring, we find that the
$n_\rho = 0.5$ case entrains and grows with depth, similarly to the Boussinesq regime.
On the other hand, the $n_\rho = 3$ case has a radius which remains approximately constant
over time, and it reaches the bottom of the domain in many fewer nondimensional freefall
time units.

\begin{figure*}[t!]
    \includegraphics[width=\columnwidth]{evolution_colormeshes.png}
    \caption{
    \label{fig:evolution_colormeshes} }
\end{figure*}

@@

We further assume that the volume-integrated momentum, $P_z \approx \rho V w_{th}$,
where $w_{th}$ is the vertical velocity of the thermal as a whole, and the volume
of the fluid region propogating with the thermal is a spheroid, $V = V_0 r^3$, for
some constant $V_0$. Plugging Eqn.~\ref{eqn:r_theory} into this approximation, we
find that
\begin{equation}
\rho^{-1/2}w = \left(\frac{(\pi\Gamma)^{3/2}}{V_0}\right)\frac{\beta B t + M_0}{(B t + I_0)^{3/2}}.
\end{equation}
Substituting $w = dz/dt$ and $\tau \equiv (Bt + I_0)/\Gamma$, We retrieve
\begin{equation}
\frac{dz}{\sqrt{\rho(z)}} 
= \left(\frac{\pi^{3/2}\Gamma}{V_0}\right)
\left(\beta\Gamma\tau^{-1/2} + [M_0 - \beta I_0]\tau^{-3/2}\right)d\tau.
\label{eqn:general_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this
result can be straightforwardly integrated with $\rho(z)$ plugged in in order to
find the position of the thermal versus time. While we will leave this result
general for now, we will plug in our polytropic stratification at the end of
section \ref{sec:experiment}, and show that the resulting expression does a
good job of explaining the evolution of thermals in these atmospheres in
section \ref{sec:results}.

@@

\subsection{Anelastic Simulations}
In this work, we primarily study the evolution of 2D, azimuthally symmetric, anelastic
thermals in cylindrical coordinates. 
We later verify that these simulations produce the same results a 3D fully compressible
simulations in 3D cartesian domains (see sec. REF). 
The LBR anelastic equations are \citep{lecoanet&all2014},
\begin{gather}
\td{\grad}\cdot\td{\bm{u}} = -\td{w}\partial_{\tilde{z}} \ln\rho_0 \\
\frac{D \td{\bm{u}}}{D \td{t}} = -\td{\grad} \td{\varpi} + g\frac{\td{S_1}}{c_P}\hat{z} + \frac{1}{\rho_0}\td{\grad}\cdot\left(\mu\td{\lilstressT}\right) \\
\frac{D \td{S_1}}{D\td{t}} = \frac{1}{\rho c_P}\td{\grad}\cdot\left(\kappa T_0 \td{\grad} \td{S_1}\right) + \frac{\mu}{\rho_0 T_0}\td{\sigma_{ij}}\partial_{\td{x_i}}\td{u_j}.
\end{gather}
where terms with tildes are dimensional, and where
$D/D\td{t} = \partial/\partial \td{t} + \td{\bm{u}}\cdot\td{\grad}$. In our azimuthally symetric
domain, we assume that $\partial_\phi = 0$; as the initial conditions of our simulations are at rest and have
no azimuthal velocity, $u_\phi$, we assume that $u_\phi = 0$; therefore $\bm{u} = u_r \hat{r} + w\hat{z}$. 
Under this approximation, the components of the stress tensor are
\begin{equation}
\begin{split}
&\td{\sigma}_{rr} = 2\frac{\partial \td{u_r}}{\partial \td{r}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}},\,\,\,\,\,\,
\td{\sigma}_{rz}     = \td{\sigma}_{zr} = \frac{\partial \td{w}}{\partial \td{r}} + \frac{\partial \td{u_r}}{\partial \td{z}}, \\
&\td{\sigma}_{\phi\phi} = 2\frac{\td{u_r}}{\td{r}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}},\,\,\,\,\,\,\,
\td{\sigma}_{r\phi}  = \td{\sigma}_{\phi r}  = 0, \\
&\td{\sigma}_{zz}       = 2\frac{\partial \td{w}}{\partial \td{z}} - \frac{2}{3}\td{\grad}\cdot\td{\bm{u}}, \,\,\,\,\,\,
\td{\sigma}_{\phi z} = \td{\sigma}_{z \phi}  = 0,\qquad
\end{split}
\end{equation}
Furthermore, we assume that the dynamic viscosity, $\mu = \rho \nu$, and the
thermal conductivity, $\kappa = \rho \chi$, are both uniform throughout the domain.
The diffusivities
$\nu$ and $\chi$ therefore scale inversely with the density.

We nondimensionalize these equations in the same manner as in
\citet{lecoanet&jeevanjee2018} such that
the length scale is the diameter of the initial thermal perturbation
and the velocity scale is the freefall velocity. The timescale is thus
the freefall crossing time of this unit length. Mathematically,
\begin{equation}
\begin{split}
\td{\grad}\rightarrow(\td{L}_{th}^{-1})\grad, \qquad&
\td{S}_1 \rightarrow(\Delta\td{S})S_1,\\
\td{\bm{u}} \rightarrow (\td{u}_{th})\bm{u}, \qquad&
\td{\varpi} \rightarrow (\td{u}_{th}^2)\varpi,\\
\partial_{\tilde{t}} \rightarrow (\td{u}_{th}/\td{L}_{th})\partial_t,\qquad&
\end{split}
\end{equation}
with
\begin{equation}
\tilde{u}_{th}^2 = \frac{g \tilde{L}_{th} \Delta \tilde{s}}{c_P}, \qquad
\text{Re}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\nu}, \qquad
\text{Pr}_{\text{ff}} = \frac{\tilde{u}_{th} \tilde{L}_{th}}{\chi}.
\end{equation}
As the diffusivities scale with depth, Re$_{\text{ff}}$ is specified at the
thermal's initial depth.
The resulting equations are,
\begin{gather}
\DivU = -w \partial_z \ln\rho_0, \\
\begin{split}
\partial_t \bm{u} + \bm{u}\cdot\grad\bm{u} =& \\
- \grad \varpi + S_1\hat{z} &
+ \frac{1}{\text{Re}_{\text{ff}}}\left[\grad^2 \bm{u} + \frac{1}{3}\grad(\DivU)\right] 
\end{split}\\
\begin{split}
\partial_t S_1 + \bm{u}\cdot\grad S_1 =& \\
\frac{1}{\text{Re}_{\text{ff}}}\left(\frac{1}{\text{Pr}_{\text{ff}}\rho_0c_P }\right.&[\grad^2 S_1 + \partial_z\ln T_0 \cdot\partial_z S_1]\\
&+ \left.\frac{-(\grad_{ad})}{\rho_0 T_0}\sigma_{ij}\partial_{x_i}u_j \right)
\end{split}
\end{gather}
where in this nondimensionalization, the adiabatic temperature gradient is defined
in the next subsection.

\subsection{Atmosphere \& Initial conditions}
We study an ideal gas whose equation of state is $P = \rho T$ and whose stratification
is a perfectly adiabatic polytrope,
\begin{gather}
T_0 = 1 + (\grad_{\text{ad}})(z - L_z) \\
\rho_0 = T_0^{m_{\text{ad}}},
\end{gather}
where $m_{\text{ad}} = (\gamma-1)^{-1}$, and the adiabatic temperature 
gradient in these nondimensional atmospheres is
$$
\grad_{\text{ad}} = -\frac{(e^{n_\rho/m_{ad}} - 1)}{L_z}\frac{m_{ad} + 1}{c_P},
$$
where $L_z$ is the atmospheric depth in units of thermal diameters.

To initialize the simulation, we specify a spherical initial specific entropy perturbation,
\begin{equation}
S_1 = - A(1 - \text{erf}\left(\frac{r' - r_{th}}{\delta}\right))/2,
\label{eqn:thermal_IC}
\end{equation}
where $A = 1$ for our scaled equations.
Here, $r' = \sqrt{r^2 + (z - z_0)^2}$, where $z_0 = L_z - 3*r_{th}$,
with the thermal radius set as $r_{th} = 0.5$, and a smoothing width, $\delta = 0.1$.

\subsection{Fully Compressible Simulations}
In order to verify the validity of our 2D Anelastic simulations, we evolve select
thermals according to the 3D Navier Stokes equations in a cartesian domain. We use
the $(T, \ln\rho)$ formulation of the equations in which we have previously studied fully compressible
convection at low and high Mach number \citep{anders&brown2017, lecoanet&all2014},
\begin{gather}
\frac{D \ln\rho}{Dt} + \DivU = 0 \\
\frac{D \bm{u}}{D t} = -\grad T - T\grad\ln\rho - g\hat{z} + \frac{1}{\rho}\grad\cdot\left(\mu\lilstressT\right) \\
\frac{D T}{Dt} + (\gamma-1)T\DivU  = \frac{1}{\rho c_V}\grad\cdot\left(\kappa \grad T\right) + \frac{\mu}{\rho c_V}\sigma_{ij}\partial_{x_i}u_j.
\end{gather}
where the viscous stress tensor is defined as
\begin{equation}
\sigma_{ij} = \left(\partial_{x_i}u_j + \partial_{x_j}u_i - \frac{2}{3}\delta_{ij}\grad\cdot\bm{u}\right).
\end{equation}
These equations are nondimensionalized on the temperature gradient length scale such that
$\grad_{\text{ad}} = -1$ and the nondimensional timescale is the sound crossing time of that
unit length.

In setting the specific entropy, $S = c_V \ln T - R^{-1}\ln\rho$, to an equivalent condition
to that specified in  Eqn.~\ref{eqn:thermal_IC}, we note that it is essential that the
initial perturbation be in pressure equilibrium. The set of initial conditions that achieves this
is
\begin{equation}
\ln\rho_1 = S_1/c_P, \qquad T_1 = T_0(e^{-\ln\rho_1} - 1).
\end{equation}
We specify the magnitude of the initial entropy perturbations as 
$A = \epsilon = 10^{-4}$, such that the mach number of the resultant thermal is O($10^{-2}$) or less.

The depth of the atmosphere in these fully compressible domains is set by the number of density
scale heights as $L_z = e^{n_\rho/m_{ad}} - 1$, and in order to compare results from these
simulations to our nondimensionalized thermal simulations, we rescale all length scales in
post-processing by $\ell = 20/L_z$, all timescales by $\tau = \sqrt{\ell \epsilon}$, and
the entropy by $s = \epsilon^{-1}$.

\subsection{Solution for thermal evolution in a Polytrope}
Here we take the general theory described in Eqn.~\ref{eqn:general_theory} and apply it
to the specific case of the polytropes we study here. 
We note that, $\rho_0 = T_0^{m_{ad}}$ and
$dT_0 = (\grad_{\text{ad}}) dz$, such that we must solve
\begin{equation}
T_0^{-m_{ad}/2} dT_0 
= \left(\frac{\pi^{3/2}\Gamma\grad_{\text{ad}}}{V_0}\right)
\left(\beta\Gamma\tau^{-1/2} + [M_0 - \beta I_0]\tau^{-3/2}\right)d\tau.
\label{eqn:general_theory}
\end{equation}
Assuming that $m_{ad} < 2$ and through the use of simple power law integrals, we
find that
\begin{equation}
T_0(t) = \left(C[\xi(\tau) - \xi(\tau_0)] + T_0(t_0)^{1/\alpha}  \right)^{\alpha},
\end{equation}
with
\begin{equation}
\xi(\tau) = \beta\Gamma\tau^{1/2} - (M_0 - \beta I_0) \tau^{-1/2},
\end{equation}
$C \equiv (2 \pi^{3/2} \Gamma (\grad_{\text{ad}}))/(\alpha V_0 B)$
